---
title: "Principal component analysis (PCA)"
#author: "Daniel Steffen Unkel, Thomas Klein-Heßling"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## PCA and PCR



The data frame `USairpollution` in the R package `HSAUR2` contains air pollution data of 41 US cities. 

```{r message=FALSE,warning=FALSE}
library(HSAUR2)
head(USairpollution)
attach(USairpollution)
```

The question of most interest about these data is probably how the pollution level as measured by sulphur dioxide concentration (`SO2`) is related to the six other variables. However, we first ignore the `SO2` variable and concentrate on the others, two of which relate to human ecology (`popul`,`manu`) and four to climate (`temp`, `wind`, `precip`, `predays`). We replace the `temp` by a variable representing negative temperatures, since then all six variables are such that high values represent a less attractive environment.

```{r}
USairpollution$negtemp <- temp*(-1)
USairpollution <- USairpollution[,c(1,8,3:7)]
attach(USairpollution)
head(USairpollution)
```

```{r}
round(cor(USairpollution[,-1]),2) # sample correlation matrix
plot(USairpollution[,-1]) # matrix scatterplot
```


#### PCA

We carry on with a PCA of the data for all 41 cities. We take the correlation matrix as an input for PCA as the six variables are measured on very different scales.

```{r}
usair.pc<-princomp(USairpollution[,-1],cor=TRUE) 
summary(usair.pc,loadings=TRUE)
```

The first three PCs all have variances greater than one 
and together account for almost 85% of the variance of the original variables. The loadings are scaled such that their sum of squares is unity (blanks indicate near zero values).

**Interpretation of PCs**

1st component: some index of "quality of life" (QoL) with high values indicating a relatively "poor" environment.

2nd component: high coefficients for `precip` and `predays`, and might be labeled as the "wet weather" component.

3rd component: Contrast between `precip` and `negtemp`, and will separate cities having high temperatures and high rainfall from those that are colder but drier. A suitable label might be simply "climate type".

**PC scores**

```{r}
usair.pc$scores[,1:3]
```

```{r}
par(pty="s")
plot(usair.pc$scores[,1],usair.pc$scores[,2],
ylim=range(usair.pc$scores[,1]),
xlab="QoL",ylab="Wet weather",type="n",lwd=2)
text(usair.pc$scores[,1],usair.pc$scores[,2],
labels=abbreviate(row.names(USairpollution)),cex=0.7,lwd=2)

par(pty="s")
plot(usair.pc$scores[,1],usair.pc$scores[,3],
ylim=range(usair.pc$scores[,1]),
xlab="QoL",ylab="Climate type",type="n",lwd=2)
text(usair.pc$scores[,1],usair.pc$scores[,3],
labels=abbreviate(row.names(USairpollution)),cex=0.7,lwd=2)

par(pty="s")
plot(usair.pc$scores[,2],usair.pc$scores[,3],
ylim=range(usair.pc$scores[,2]),
xlab="Wet weather",ylab="Climate type",type="n",lwd=2)
text(usair.pc$scores[,2],usair.pc$scores[,3],
labels=abbreviate(row.names(USairpollution)),cex=0.7,lwd=2)
```
